---
title: "CHEP898_A3:Missing Data"
author: "Lina"
date: "2025-01-29"
output: 
  html_document:
      keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(tidyverse)
library(visdat)
library(naniar)
library(mice)
```

```{r}
#  R function that counts how many columns in your dataset belong to each type 
#(col_logical(), col_double(), col_character()):
count_column_types <- function(data) {
  # Count the number of columns for each type
  counts <- data %>%
    summarise(
      logical_cols = sum(sapply(., is.logical)),
      double_cols = sum(sapply(., is.double)),
      character_cols = sum(sapply(., is.character))
    )
  
  # Return the counts as a named vector
  return(counts)
}

```

# Assignment Introduction

**Assignment Goal:** Demonstrate an understanding of various imputation techniques and their implications for analysis.

**Assignment Objectives:**

- Understand the nature and patterns of missing data in a large dataset.

- Apply different imputation methods and assess their suitability for the dataset.

- Compare the results of different imputation techniques using statistical and visual methods.

- Develop insights into the impact of missing data handling on subsequent analyses.

# Dataset Exploration

## Load the provided health administrative dataset.
```{r}
# Load the data
data <- read_csv("can_path_data.csv")

# Select all columns between ID and HS_GEN_HEALTH, also select column PA_TOTAL_SHORT, 
# PM_BMI_SR, all columns that contain the string "_EVER" anywhere in their names,
# and all columns that contain the string "WRK_" anywhere in their names
select_data <- data %>% select(ID:HS_GEN_HEALTH, PA_TOTAL_SHORT, PM_BMI_SR, 
                        contains("_EVER"), contains("WRK_"))

# Check dimensions and missing data
cat("The select data has ", ncol(select_data), "columns, ", nrow(select_data), "rows, and ",
    sum(is.na(select_data)), "missing values.\n")
```

## Explore the dataset to identify the extent, patterns, and potential reasons for missing data.

```{r}
missing_table <- miss_var_summary(select_data)

# Count the number of variables with pct_miss > 85%
count_high_missing <- missing_table %>%
  filter(pct_miss > 85) %>%
  summarise(count = n())

# Display the result
cat("Number of variables with missing rate more than 85%:", count_high_missing$count, "\n")

```

```{r}
# Extract variable names with more than 85% missing data
high_missing_vars <- missing_table %>%
  filter(pct_miss > 85) %>%
  pull(variable)

high_missing_data <- select_data %>%
  select(all_of(high_missing_vars))
```

Check how many columns in the data read as logical variable.
```{r}
# Check how many columns in the high missing data belong to each type 
#(col_logical(), col_double(), col_character())
column_type_counts_high <- count_column_types(high_missing_data)

cat("Number of logical variables in high missing data:", column_type_counts_high$logical_cols, 
    ", and number of numeric variables in high missing data:", column_type_counts_high$double_cols, "\n")

# Get rid the columns read as logical
logical_variable_names <- high_missing_data %>%
  select_if(is.logical) %>%
  colnames()

select_data <- select_data %>% select(!(logical_variable_names))

```



## Summarize findings using tables, charts, or heatmaps to visualize missingness.

```{r}
# Missing data table
missing_table

# Visualize missing data
## The bar plot shows how many rows in the dataset have missing values for each variable
gg_miss_var(high_missing_data) + ggtitle("Missing Data by Variable")

## UpSet is used to display the intersections between multiple sets of data
gg_miss_upset(high_missing_data)
```

From the table above, we can see that there are 102 variables with more than 85% missing data. Among these 102 variables, 10 are logical variables, and 99 are numeric variables. The logical variables should be excluded, and since the 99 numeric variables are difficult to explain, we do not include them in our further data analysis. This is also supported by the following plots.

The bar chart provides a general overview of missing data per variable. It shows that all 102 variables have more than 35,000 missing data points.

The UpSet plot illustrates patterns of missing data across variables. The highest bar (the leftmost one) represents 41,127 rows where certain variables have missing data. The middle bar corresponds to 44 rows where two specific variables have missing values simultaneously, and the rightmost bar corresponds to 16 rows where another subset of variables is missing together. The large number of rows represented by the 41,127-row bar in the UpSet plot indicates that many variables have missing data for a significant portion of the dataset, which could affect the analysis.


```{r}
select_data_less_85 <- select_data %>%
          select(where(~sum(is.na(.x))/length(.x) < 0.85))

# Double check the data
missing_table_less_85 <- miss_var_summary(select_data_less_85)
missing_table_less_85
```

At this step, we can see that all the missing rates are less than 85%.

# Apply Imputation Methods

## Select and apply at least three different methods for imputing missing data, such as:

  - Mean/Median/Mode imputation.\
  - K-Nearest Neighbors (KNN) imputation.\
  - Multiple Imputation by Chained Equations (MICE).
  
## Document the implementation process for each method.

# Evaluation of Imputation Methods
## Compare the imputed datasets by analyzing:

  - Changes in key summary statistics (e.g., means, variances).
  - The preservation of relationships between variables (e.g., correlations).
  - Visual comparisons of distributions before and after imputation.
  
# Analysis of Imputed Data
## Conduct a simple statistical analysis on the imputed datasets to illustrate the downstream effects of different imputation methods.

##Discuss how the choice of imputation method impacts the analysis results.

# Interpretation and Reporting
## Provide a detailed comparison of the methods, discussing their strengths, weaknesses, and suitability for the dataset.
## Reflect on the challenges of handling missing data in health research.



